Dancing on Water: The Choreography of Sulfur Dioxide Adsorption to Aqueous Surfaces

### 2011-07-01 Article.tex
Fixed bibliography note issue, and made title page w/ abstract.
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

\title{Sulfur Dioxide on Water's Surface}
\author{Eric S. Shamay \and Kevin E. Johnson \and Geraldine L. Richmond}

@@

%\maketitle

@@

\begin{abstract}
Sulfur dioxide binding to water surfaces is a model system for the study of gaseous adsorption processes on aqueous systems. Equilibrium and adsorbing \suldiox~molecules in a pure water interfacial system are modeled by classical molecular dynamics simulations using fully polarizable models for both \suldiox~and \wat. Neat-water systems with a single \suldiox, and high \suldiox~concentration saturated water systems are modeled at equilibrium and using steered molecular dynamics (SMD) to ascertain the orientational profile of a transiting \suldiox~molecule. Orientational depth profiling reveals the preferred orientations of \suldiox~during adsorption, and the complementary behaviors of surface waters adsorbing the gaseous solute.
\end{abstract}

>>>>

\title{Dancing on Water: The Choreography of Sulfur Dioxide Adsorption to Aqueous Surfaces}
\author{Eric S. Shamay \and Kevin E. Johnson \and Geraldine L. Richmond}

@@

\maketitle

@@

\begin{abstract}
One might expect the high surface tension of water to be a barrier to adsorption of a gas into the liquid phase, but we know that gaseous adsorption on and into a water surface is a common phenomenon on this planet.  What is not commonly known is how an atmospheric gas such as \suldiox~and molecules at the water surface can overcome the barrier created by strong water-water surface bonding interactions.  What this interplay looks like, the distances from the water surface at which these attractive interactions begin, and how they influence the orientational nature of both \suldiox~and surface water molecules is the focus of this computational study.  The results fill a void in the information about this system existing from previous experimental studies by providing information about the dimensional nature of the gas-surface interactions, and the details of how the two species twist and turn orientationally with increased surface interactions.  Classical molecular dynamics have been employed in both equilibrium and steered molecular dynamics (SMD) simulations for \suldiox~at a neat-water surface, and at a surface with high interfacial \suldiox~concentrations.  The results provide new molecular insights for understanding the interaction of this prevalent gas on aerosols and other aqueous surfaces in the environment.
\end{abstract}

### 2011-07-01 Conclusions.tex
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

Through these simulations we are able to characterize the orientational behavior of \suldiox~during and after adsorption. Our equilibrium simulations show that a single \suldiox~molecule, representing a low concentration, has a high surface affinity. At a high \suldiox~concentration, unreacted \suldiox~molecules are found further out of the water phase as a bound layer of \suldiox~crowds the surface and binds all the surface waters. The orientation of \suldiox~on the water surface was found to be similar for both low and high concentrations. Those \suldiox~molecules at or below the surface strongly orient with the sulfur atom pointed in towards the water bulk, and the oxygen atoms out towards the gas phase. The \suldiox~slightly above the water surface loses the orientational preference within 10 $\AA$ and those further from the water are more isotropically oriented. Figure \ref{fig:so2-surface-cartoon} depicts what the neat-water and saturated surfaces might look like for both \suldiox~and \wat~orientations and locations.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~has very similar orientation to those already bound to the water surface. The \suldiox~reorients as it makes its first contact with the water surface. Within 5 $\AA$ of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

Having now examined this passageway of \suldiox~from the gas to adsorbed aqueous phases, we may turn to further filling in the details of adsorption and interfacial chemistry of gas molecules. What species form during the adsorption transition, and how do molecular properties affect the process? Can we form a theory that will more fully explain the transit of many small molecules of environmental and industrial interest? This study is one of several aiming to characterize \suldiox~adsorption and behavior on aqueous surfaces. We plan to report further results of ongoing computational simulation and experimental studies regarding temperature and chemical constituent effects on atmospherically relevant water surfaces.

>>>>

Through these simulations we characterize the orientational behavior of \suldiox~during and after adsorption. Our equilibrium neat-water simulations show that a single \suldiox~molecule, representing a low concentration, has a high surface affinity. At a high \suldiox~concentration in the saturated systems, \suldiox~molecules are also surface active, and are found further out of the water phase than at the lower concentration. These \suldiox~form a bound layer of molecules that crowd the surface and interact with the surface waters. The orientation of \suldiox~on the water surface was found to be similar for both low and high concentrations. Those \suldiox~molecules at or below the surface strongly orient with the sulfur atom pointed in towards the water bulk, and the oxygen atoms out towards the gas phase. The \suldiox~slightly above the water surface loses the orientational preference within 10 $\AA$ and those further from the water are more isotropically oriented. Figure \ref{fig:so2-surface-cartoon} depicts what the neat-water and saturated surfaces might look like for both \suldiox~and \wat~orientations and locations.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~has very similar orientation to those already bound to the water surface. The \suldiox~reorients as it makes its first contact with the water interface. Within 5 $\AA$ of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

Having now examined the transition of \suldiox~from the gas to adsorbed aqueous phases, we are better able to research the many questions remaining of adsorption and interfacial chemistry of gas molecules. What species form during the adsorption transition, and how do molecular properties affect the process? Can we form a theory that will more fully explain the transit of many small molecules of environmental and industrial interest? This study is one of several characterizing \suldiox~adsorption and behavior on aqueous surfaces. We plan to report further results of ongoing computational simulation and experimental studies regarding temperature and chemical constituent effects on atmospherically relevant water surfaces.

### 2011-07-01 DensityResults.tex
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

In the simulated neat-water slab, the result that the \suldiox~molecular density distribution concentrates near the water surface indicates an affinity for the interfacial region. During the course of the simulation, the \suldiox~does not venture into the bulk water nor does it escape the water surface into the gas phase, but remains located within 5-\angs~of the surface region. This is consistent with what is found experimentally in our laboratory using VSFS,\cite{Tarbuck2005,Tarbuck2006} and also supported by the computational simulations and spectral calculations of Baer et al.\cite{Baer2010} The experimental studies indicated that upon exposure of \suldiox~to a \wat~surface, a layer of solvated \suldiox~forms, modifying the structure of water above the Gibbs dividing surface (GDS). In the experimental system, the \suldiox~bonding interaction with the free OH oscillators is manifested in a red-shift of the free OH frequency indicative of a bonding interaction.

The saturated solution simulation results indicate that under the high concentration of \suldiox, the \suldiox~accumulation at the surface is increased. However, unlike the neat-water slab, the saturated slab has a non-zero bulk concentration of \suldiox. The added concentration of \suldiox~creates a layer of molecules bound to the top of the water surface. The center of the \suldiox~density distribution is further into the gas phase than for the neat-water surface with a single-\suldiox~molecule. Because of the limitations of the classical model in accurately reproducing the first hydration shell around the \suldiox,\cite{Baer2010} we do not draw specific conclusions about the hydrate complex's geometry. However, the surface affinity of the hydrated \suldiox~is well reproduced by both the classical and ab initio methods.\cite{Baer2010}

>>>>

In the simulated neat-water slab, the \suldiox~molecular density distribution concentrates near the water surface location, indicating an affinity for the interfacial region. During the course of the simulation, the \suldiox~does not venture into the bulk water nor does it escape the water surface into the gas phase, but remains located within 5-\angs~of the surface region. This is consistent with what is found experimentally in our laboratory using VSFS,\cite{Tarbuck2005,Tarbuck2006} and also supported by the computational simulations and spectral calculations of Baer et al.\cite{Baer2010} The experimental studies indicated that upon exposure of \suldiox~to a \wat~surface, a layer of solvated \suldiox~forms, modifying the structure of water in the upper surface region. In the experimental system, the \suldiox~bonding interaction with the free OH oscillators is manifested in a red-shift of the free OH frequency indicative of a bonding interaction.

The saturated solution simulation results indicate that under the high concentration of \suldiox, the \suldiox~accumulation at the surface is increased. However, unlike the neat-water slab, the saturated slab has a non-zero bulk concentration of \suldiox. The added concentration of \suldiox~creates a layer of molecules bound to the top of the water surface. The center of the \suldiox~density distribution is further into the gas phase than for the neat-water surface with a single-\suldiox~molecule. Because of the limitations of the classical model in accurately reproducing the first hydration shell around the \suldiox,\cite{Baer2010} we do not draw specific conclusions about the hydrated surface complex's geometry. However, the surface affinity of the hydrated \suldiox~is well reproduced by both the classical and ab initio methods.

### 2011-07-01 EquilibratedOrientation.tex
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

The orientation depth-profiles for \wat~are shown in Figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. The interfacial region for both these calculations and the VSF experiments is defined as the region where mlecular orientational anisotropy exists around the surface water location. Referring to Figure \ref{fig:water-orientation}, calculations indicate an interfacial width of approx. 10\angs~for the neat-water system, and approx. 16 \angs~for the saturated system. In both systems the strongest orientational preference is found at the slab surfaces (distances greater than 0\angs) where the water is furthest towards the gas phase. Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The narrow "peninsula" of high intensity that extends in the $\cos(\theta)$ plots to the right of the -5\angs~location shows the overall preference of water to orient at the surface. The $\cos(\phi)$ plots are similar to each other with a narrow region of reorientation, but the effect in the interfacial region is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change.

@@

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the water orientation of the waters furthest into the gas phase. For the saturated solution the resulting orientation of waters above 0\angs, shown in the saturated plot of $\cos(\theta)$ of Figure \ref{fig:water-orientation}, is with a bisector pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced further from the water phase, and above 5\angs~the $\cos(\theta)$ distribution is completely centered around $\cos(\theta)=+1$ (see Figure \ref{fig:water-angles}).

@@

In the interfacial region Within 0-5\angs~of the water surface location (the approximate distance from the surface that \suldiox~begins interacting with the outer-most waters) the angular distribution of the single \suldiox~(in the neat-water system) is concentrated primarily in $\cos(\theta)>0$. The peak of the distribution occurs at $\cos(\theta)=1$. This indicates that the \suldiox~bisector points out of the water surface, with the sulfur atom pointing towards the aqueous bulk, and the two oxygens pointing into the gas phase. This same distribution occurs in the saturated system for positions below 5\angs. Beyond 5\angs~above the surface both distributions become isotropic. Promixity to the water highly orients the \suldiox~with the sulfur atom pointing in towards the water bulk. Moving further away from the water surface, and interacting less with \wat~molecules allows for greater orientational freedom as exhibited in the isotropy of the distributions above 5\angs. %Because both the neat-water and saturated systems show rather similar distributions, it is possible that the concentration of \suldiox~does not strongly affect the orientation of the \suldiox, unlike the orientations of waters.

>>>>

The orientation depth-profiles for \wat~are shown in Figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. The interfacial region for both these calculations and the VSF experiments is defined as the region where molecular orientational anisotropy exists around the surface water location. Referring to Figure \ref{fig:water-orientation}, calculations indicate an interfacial width of approximately 10\angs~for the neat-water system, and approximately 16 \angs~for the saturated system. In both systems the strongest orientational preference is found at the slab surfaces (positions above 0\angs) where the water is furthest towards the gas phase. Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The narrow "peninsula" of high intensity that extends in the $\cos(\theta)$ plots to the right of the -5\angs~location shows the overall preference of water to orient at the surface. The $\cos(\phi)$ plots are similar to each other with a narrow region of reorientation, but the effect in the interfacial region is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change.

@@

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the orientation of those waters furthest into the gas phase. For the saturated solution the resulting orientation of waters above 0\angs, shown in the saturated plot of $\cos(\theta)$ of Figure \ref{fig:water-orientation}, is with a bisector pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced further from the water phase, and above 5\angs~the $\cos(\theta)$ distribution is completely centered around $\cos(\theta)=+1$ (see Figure \ref{fig:water-angles}).

@@

In the interfacial region within 0-5\angs~of the water surface location (the approximate distance from the surface that \suldiox~begins interacting with the outer-most waters) the angular distribution of the single \suldiox~(in the neat-water system) is concentrated primarily in $\cos(\theta)>0$. The peak of the distribution occurs at $\cos(\theta)=1$. This indicates that the \suldiox~bisector points out of the water surface, with the sulfur atom pointing towards the aqueous bulk, and the two oxygens pointing into the gas phase. This same distribution occurs in the saturated system for positions below 5\angs. Beyond 5\angs~above the surface both distributions become isotropic. Promixity to the water highly orients the \suldiox~with the sulfur atom pointing in towards the water bulk. Moving further away from the water surface, and interacting less with \wat~molecules allows for greater orientational freedom as exhibited in the isotropy of the distributions above 5\angs. %Because both the neat-water and saturated systems show rather similar distributions, it is possible that the concentration of \suldiox~does not strongly affect the orientation of the \suldiox, unlike the orientations of waters.

### 2011-07-01 Introduction.tex
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of a molecule between the two bulk phases. What is the interplay between the gas and surface water molecules and when does one begin to influence the behavior of the other? What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  Is molecular orientation of either the gas or surface molecules a factor in the adsorption process? Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the full resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site. This type of information can be determined computationally, and when coupled to the experimental studies can provide a more comprehensive picture of the gas-liquid surface adsorption process.

An important gas for developing a picture of gaseous adsorption and entry into a wter surface is \suldiox.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011} Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.

In this work we provide a molecular picture of \suldiox~adsorption on a water surface.  The studies demonstrate the strong orientational effect that surface water molecules have on the adsorbing gas during the approach and entry into the surface region, and how surface water molecules rise towards the gas phase to assist in the capture of \suldiox~molecules at both high and low \suldiox~surface concentrations.

These computational studies complement and significantly expand the picture developed in our recent experimental vibrational sum frequency spectroscopy (VSFS) studies of \suldiox~adsorption of aqueous solutions of various compositions and temperatures, and the subsequent studies by Baer et al.\cite{Baer2010} using both classical and ab initio simulations.  The most recent experimental studies have shown that whereas the binding of gaseous \suldiox~to a water surface is greatly enhanced at cold temperatures, the reversibility of the adsorption process remains.  Complimentary experiments showed that low pH aqueous environments inhibit the bulk reactions of \suldiox, but do not affect the surface binding or its reversibility.  What is apparent in the VSF spectra obtained in all of these experiments is the tendency of water to reorient upon surface bonding, with the effect becoming more pronounced at high \suldiox~surface concentrations. Since the \suldiox~molecule was not specifically probed, conclusions on how \suldiox~bonding contributes to reorienting surface water molecules and the orientation of \suldiox~itself upon approach and surface bonding could only be inferred. Obtaining a more complete picture of this adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

Our recent vibrational sum frequency spectroscopy (VSFS) experiments have provided insight into the adsorption and reaction of gases in atmospheric aerosols.\cite{Tarbuck2005,Tarbuck2006} These studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. A computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.

>>>>

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of a molecule between the two bulk phases. What is the interplay between the gas and surface water molecules, and when does one begin to influence the behavior of the other? What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  Is molecular orientation of either the gas or surface molecules a factor in the adsorption process? Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the full resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site. This type of information can be determined computationally, and when coupled to the experimental studies can provide a more comprehensive picture of the gas-liquid surface adsorption process.

An important gas for developing a picture of gaseous adsorption and entry into a water surface is \suldiox.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011} Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.

In this work we provide a molecular picture of \suldiox~adsorption on a water surface.  This study demonstrates the strong orientational effect of surface water molecules on adsorbing gas during the approach and entry into the surface region at both high and low \suldiox~surface concentrations.  Obtaining a more complete picture of this adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

These computational studies complement and significantly expand the picture developed in our recent experimental vibrational sum frequency spectroscopy (VSFS) studies of \suldiox~adsorption of aqueous solutions of various compositions and temperatures,\cite{Tarbuck2005,Tarbuck2006} and the subsequent studies using both classical and ab initio simulations.  These experimental studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. The computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.  The most recent experimental studies have shown that whereas the binding of gaseous \suldiox~to a water surface is greatly enhanced at cold temperatures, the reversibility of the adsorption process remains.  Complimentary experiments showed that low pH aqueous environments inhibit the bulk reactions of \suldiox, but do not affect the surface binding or its reversibility.  What is apparent in the VSF spectra obtained in all of these experiments is the tendency of water to reorient upon surface bonding, with the effect becoming more pronounced at high \suldiox~surface concentrations. Since the \suldiox~molecule was not specifically probed, conclusions on how \suldiox~bonding contributes to reorienting surface water molecules and the orientation of \suldiox~itself upon approach and surface bonding could only be inferred.

### 2011-07-01 Methods.tex
Lots of small corrections (typos!!) to draft 8 handed back for polishing.

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the equilibrated neat-water and saturated systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfully model chemical events.\cite{Isralewitz2001,Giorgino2011,Bizzarri2011,Strzelecki2009,Patargias2009,Liu2006} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a time step of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

	The first portion of the studies involved creating orientational depth-profiles in the interfacial region comprised of \suldiox~and \wat~upon exposure and adsorption of the gas. To do this we must define a reference point in thie interfacial region which we refer to here as the water surface location. Recognizing that a liquid surface is a dynamic boundary that is neither flat nor stationary. Several previous studies have used the technique of fitting a line shape to the averaged density profile of the water, and extracting interfacial shape and location parameters to define the water surface location.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given time step. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes. The analysis presented here attempts to retain these subtleties through the use of a ``corrected'' coordinate system.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin grey line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, 70 water molecules were used for calculating instantaneous water surface locations.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(grey) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit using a hyperbolic tangent function (black) to extract location and width data for the water surface. Position values are based on the atomic position data of the water oxygens and \suldiox~sulfurs from the raw molecular dynamics output. (B) The instantaneous location of the outer \wat~monolayer was calculated for each simulation time step (blue), and the surface location extracted from the density fitting was also shown for reference (horizontal dashed line). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}

>>>>

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the neat-water and saturated starting systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were then evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfully model chemical events.\cite{Isralewitz2001,Giorgino2011,Bizzarri2011,Strzelecki2009,Patargias2009,Liu2006} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a time step of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

	The first portion of the studies involved creating orientational depth-profiles in the interfacial region comprised of \suldiox~and \wat~upon exposure and adsorption of the gas. Recognizing that a liquid surface is a dynamic boundary that is neither flat nor stationary, we must define a reference point in the interfacial region which we refer to here as the water surface location. Several previous studies have used the technique of fitting a line shape to the averaged density profile of the water, and extracting interfacial shape and location parameters to define the water surface location.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given time step. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes. The analysis presented here attempts to retain these subtleties through the use of a ``corrected'' coordinate system.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin gray line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(Gray) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit using a hyperbolic tangent function (black) to extract location and width data for the water surface. Position values are based on the atomic position data of the water oxygens and \suldiox~sulfurs from the raw molecular dynamics output. (B) The instantaneous location of the outer \wat~monolayer was calculated for each simulation time step (blue), and the surface location extracted from the density fitting was also shown for reference (horizontal dashed line). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}
